# Installation instructions

## System requirements

The partitioning steps are currently performed using a system call to the Markov Cluster (MCL) algorithm that presently limits the use of DBF-MCL to unix-like platforms. Importantly, the `mcl` command should be in your PATH and reachable from within R.

## Step 1 - Installation of MCL

### Installation of MCL using install_mcl()

The `install_mcl()` has been developed to ease MCL installation. This function should be call automatically from within R when calling the `gene_clustering()` function. If `install_mcl()` does not detect MCL in the PATH it will install it in `~/.scigenex`.


### Installation of MCL from source

One also can install MCL from source using the following code.

```{bash , eval=FALSE, echo=TRUE}
# Download the latest version of mcl 
wget http://micans.org/mcl/src/mcl-latest.tar.gz
# Uncompress and install mcl
tar xvfz mcl-latest.tar.gz
cd mcl-xx-xxx
./configure
make
sudo make install
# You should get mcl in your path
mcl -h
```

### Installation of MCL using conda

Or you may install MCL using conda. 

```{bash , eval=FALSE, echo=TRUE}
conda install -c bioconda mcl
```

## Step 2 - Installation of SciGeneX

### From R

The scigenex library is currently not available in CRAN or Bioc. To install from github, use:

```{r, eval=FALSE, echo=TRUE}
 devtools::install_github("dputhier/scigenex")
 library(scigenex)
```

### From the terminal

Download the *tar.gz* from github. Uncompress and run the following command:

```{bash , eval=F, echo=T}
R CMD INSTALL scigenex
```

Then load the library from within R.

```{r, eval=F, echo=T}
library(scigenex)
```

