# Installation instructions

## Step 1 - Installation of MCL

You may skip this step as the latest versions of SciGeneX will call `scigenex::install_mcl()`to install MCL in `~/.scigenex` directory if this program is not found in the PATH.


### Installation of MCL from sources

Alternatively you may install MCL from sources and add it to the PATH.

```{bash , eval=F, echo=T}
# Download the latest version of mcl 
wget http://micans.org/mcl/src/mcl-latest.tar.gz
# Uncompress and install mcl
tar xvfz mcl-latest.tar.gz
cd mcl-xx-xxx
./configure
make
sudo make install
# You should get mcl in your path
mcl -h
```

### Installation of MCL from sources

Finally you may install MCL using conda. Importantly, the mcl command should be available in your PATH from within R.

```{bash , eval=F, echo=T}
conda install -c bioconda mcl
```


## Step 2 - Installation of SciGeneX

### From R

The scigenex library is currently not available in CRAN or Bioc. To install from github, use:

```{r , eval=F, echo=T}
 devtools::install_github("dputhier/scigenex")
 library(scigenex)
```

### From the terminal

Go inside the git directory and simply write:

```{bash , eval=F, echo=T}
R CMD INSTALL .
```

Next load the library from R.

```{r, eval=F, echo=T}
library(scigenex)
```

